Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-09 manuscript.qmd
add Jan's comments

@@

@@

Similarly, ipsilateral involvement correlates with contralateral metastasis. The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease. This, too, may therefore be considered a surrogate for the duration of the disease. In rare cases, bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side. <!-- citation needed? -->

Lastly, the right panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intutitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. Although these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.

@@

Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019]. We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions. We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's hidden binary RV may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some for of tumor presence (i.e., occult or clinical).

Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.

@@

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-trans-matrix}
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. This effectively masks large parts of the transition matrix to be zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-obs-matrix}
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

Note that the transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network. An example of such a DAG is shown in @fig-small-graph.

Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$. At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$. We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.

However, the remaining issue is that the value of $t_D$ is unknown, i.e. over how many time-steps the HMM should be evolved. We solve this problem by marginalizing over the time of diagnosis. Different distributions over the diagnosis times can then be choosen based on T-category. For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$. This gives us for example

@@

For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:

Where the $k$-th row in this matrix corresponds to the distribution over hidden states after $t=k-1$ time-steps.

@@

A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral. During training, one could even enforce some shared parameters between these two models, like the parameterization of the distributions over diagnose times or the spread among the LNLs. Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor.

However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement. This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.

Thus, we attempt to extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously. To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

@@

The likelihood of the diagnoses given a hidden state simply factorise: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$. And the two factors are contained in the observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.

The term representing the model's prior probability of hidden involvement does not factorize. However, if we assume that lymphatic spread typically does not cross the mid-sagittal line, we can write it as a factorising sum:

@@

This assumption makes intuitive sense: The two sides of the lymphatic network in a typical patient are approximately mirror images of each other. Thus, no major vessels cross the mid-sagittal line. There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.

Using this assumption along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

@@

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters. I.e., the ipsi- and contralateral spread rates are entirely different. However, using two sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:

1. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
2. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

@@

which is less than double in size, compared to the unilateral model. From these parameters, we now build three different transition matrices: The unchanged $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.

@@

To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.

The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by

@@

In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.

@@

The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line. But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing. We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$. We start from $t=0$, where we know that all LNLs are healthy - i.e., the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$ - and the tumor is lateralized ($\epsilon=\texttt{False}$):

$$
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
$$

Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:

@@

In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps. To keep it manageable to interpret, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.

# Methods {#sec-methods}

@@

# Results {#sec-results}

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.

@@

params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
(
  params_table.style
  .format("{:.2%}", subset=["Mean"])
  .format("± {:.2%}", subset=["Std. Dev."])
  .apply(right_align, subset=["Mean", "Std. Dev."])
  .hide()
)
```

## Prevalence Predictions

We want to investigate whether and to what extent the model can fulfill the requirements laid out in @sec-requirements. To that end, we compare its predictions for contralateral involvement against observations in the data. This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}

@Fig-model-prevalences-overall shows nicely how the model is capable of accurately taking the most important risk factors, i.e. T-category and midline extension, into account. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% <!-- TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.

However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement. This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors.

@@

## Prediction of Risk for Occult Disease

In this section, we investigate how the model may be applied clinically: We want to estimate the risk for occult disease in some or all LNLs, given the patient's individual diagnosis. In terms of our model, this diagnosis consists of the T-category, the lateralization of the tumor (does it extend over the mid-sagittal line?), and which LNLs are clinically involved, e.g. because some lymph nodes appear enlarged in an MRI scan or show increased glycolytic activity on an FDG PET/CT scan.

>>>>

Pathological involvement is thus available only for a subset of the patients and only for the dissected levels. For the remaining patients, only clinical involvement is available. For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. In brief, pathology was considered the gold standard if available, overwriting possibly conflicting reports on clinical involvement. If pathology was not available, clinical involvement determined the consensus decision. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. The Data-in-Brief publications contain a detailed description of the datasets and the data format. However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.

@@

Similarly, ipsilateral involvement correlates with contralateral metastasis. The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease. This, too, may therefore be considered a surrogate for the duration of the disease. In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side. <!-- citation needed? -->

Lastly, the right panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, figure xxx indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors . The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II. This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.

@@

@@

This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper]. In this section we provide a brief recap of the unilateral model to introduce the notation needed to extend the framework to a bilateral model describing both ipsilateral and contralateral lymph node involvement in @sec-ext-to-contra. For further details in the ipsilateral model, the reader is referred to the earlier publications [@ludwig_hidden_2021, cite 2024 graph extension paper]

<!-- 
Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019]. We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions. We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
-->

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected). Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.

@@

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-trans-matrix}
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. Consequently, many elements of the transition matrix are zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-obs-matrix}
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next. In this paper, we build on the DAG shown in @fig-small-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].

Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$. For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:

Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.

At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$. We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result. However, the remaining issue is that the value of $t_D$ is unknown, i.e. over how many time-steps the HMM should be evolved. We solve this problem by marginalizing over the time of diagnosis. Different distributions over the diagnosis times can then be choosen based on T-category. For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$. This gives us for example

@@

@@

A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral. During training, one could enforce that some parameters are shared between these two models, e.g. the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}). However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}. Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.

<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III. -->

<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->

Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time. To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

@@

The probability of the diagnoses given a hidden state factorises: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$, and the two factors are described through observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.

The term representing the model's prior probability of hidden involvement does not factorize. However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No major lymph vessels cross the mid-sagittal plane. In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e. tumor spread to the contralateral side is due to the primary tumor alone. We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:

@@

This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline. However, both sides are coupled via time. Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.

<!--There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->

Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

@@

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters. I.e., the ipsi- and contralateral spread rates are entirely different. However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:

1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-small-graph}.
2. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha \in \[0,1\]$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

@@

which is less than double in size, compared to the unilateral model. From these parameters, we now build three different transition matrices: The unchanged matrix $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.

@@

It can be assumed that most tumors that cross the midline at the time of diagnose have started as lateralized tumors and grew over the midline at a later time step. Hence, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ only applies for a subset of the time steps. 
Therefore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. A tumor starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the midsaggital plane. The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by

@@

@@

The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line. But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mixing in @eq-mixing. We can achieve the correct marginalization by iteratively building the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$. We start from $t=0$, where we know that all LNLs are healthy (the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):

$$
P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
$$

whereas all other states have zero probability. Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:

@@

# Computational methods {#sec-methods}

@@

# Results: Model evaluation {#sec-results}

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model. Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_{adv}$, we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [] for the unilateral model. The small differences may be mostly due to the differences in training dataset. Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement. We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.

@@

params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
(
  params_table.style
  .format("{:.2%}", subset=["Mean"])
  .format("± {:.2%}", subset=["Std. Dev."])
  .apply(right_align, subset=["Mean", "Std. Dev."])
  .hide()
)
```
## Illustration of the model
In this subsection we illustrate parts of the mathematical framework defined in the previous section. In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$. By parameter choice, early T-category tumors are diagnosed on average after 3 time steps. Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_{adv}=0.45)$, a parameter learned from the data to describe the higher involvement for advanced T-category tumors. The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line). The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset. The bottom panel in @fig-model-midext-evo shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right, \epsilon)$, which is illustrated in @fig-model-state-dist. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx. @fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced??? T-category tumors. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

## Prevalence predictions for contralateral involvement

The bilateral model was designed to fulfil the requirements laid out in @sec-requirements. Now, we investigate to what extent the model can quantitatively describe the patterns of lymph node involvement observed in the dataset. To that end, we compare the model's predictions for contralateral involvement against observations in the data. This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig-model-prevalences-overall.svg){#fig-model-prevalences-overall}

@Fig-model-prevalences-overall shows nicely that the model is capable of accurately accounting for the most important risk factors, i.e. T-category and midline extension. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from below 8% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 40% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line. Similarly, the prevalence of contralateral LNL III involvement increases from around 2% <!--  TODO: make this dynamic --> for early T-category lateralized tumors to almost 15% <!-- TODO: make this dynamic --> when the tumor is of advanced T-category and crosses the mid-sagittal line.

<!--  However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement. This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors. -->

@@

# Results: Prediction of Risk for Occult Disease

For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement. We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis. In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA). For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007]. For FNA we assume a specificity of 98% and a sensitivity of 80%. This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e. there are almost no false positive diagnoses.
